---
title: "R Notebook"
output: html_notebook
---
# 4.1 Conditional expressions  

Conditional expressions are one of the basic features of programming. They are used for what is calles _flow control_. The most common conditional expression is the if-else statement. In R, we can actuall perform quite a bit of data analysis without conditionals. However, they do come up occasionally, and you will need them once you start writing your own functions and packages.

Here is a very simple example showing the general structure of an if-else statement. The basic idea is to print the reciprocal of
>a
unless
> a
is 0:  

```{r}
a <- 0

if(a!=0){
  print(1/a)
} else{
  print("No reciprocal for 0.")
}
```

Let's look at one more example using the US murders data frame:
```{r}
library(dslabs)
data(murders)
murder_rate <- murders$total / murders$population*100000
```

Here is a very simple example that tells us which states, if any, have a murder rate lower then 0.5 per 100,000. The >if
statement protects us from the case in which no state satisfies the condition.
```{r}
ind <- which.min(murder_rate)

if(murder_rate[ind] < 0.5){
  print(murders$state[ind])
} else{
  print("No state has murder rate that low")
}
```

If we try it again with a rate of 0.25, we get a different answer:
```{r}
if(murder_rate[ind] < 0.25){
  print(murders$state[ind])
} else{
  print("No state has a murder rate that low.")
}
```

A related function that is very useful is
>ifelse
. This function takes three arguments: a logical and two possible answers. If the logical is
>TRUE
, the value in the second argument is returned and if
>FALSE
, the value in the third argument is returned. Here is an example:

```{r}
a <- 0
ifelse(a>0, 1/a, NA)
```

The function is particularly useful because it works on vectors. It examines each entry of the logical vector and returns elements from the vector provided in the second argument, if the entry is
>TRUE
, or elements from the vector provided in the third argument, if the entry is
>FALSE.

```{r}
a <- c(0,1,2,-4,5)
result <- ifelse(a>0, 1/a, NA)
```

This table helps us see what happened:  

 **a is_a_positive answer1 answer2 result**

?? Table

Here is an example of how this function can be readily used to replace all the missing values in a vector with zeros:
```{r}
data(na_example)
no_nas <- ifelse(is.na(na_example), 0, na_example)
sum(is.na(no_nas))
```

Two other useful functions are
>any
and
>all
. The
>any  
function takes a vector of logicals and returns
>TRUE
if any of the entries is 
>TRUE
. The 
>all
function takes a vector of logicals and returns
>TRUE 
if all of the entries are 
>TRUE
. Here is an example:

```{r}
z <- c(TRUE, TRUE, FALSE)
any(z)

all(z)

```

# 4.2 Defining functions  

As you become more experienced, you will find yourself needing to perform the same operations over and over. A simple example is computing averages. We can compute the average of a vector x using the sum and length functions: sum(x)/length(x).   Because we do this repeatedly, it is much more efficient to write a function that performs this operation. This particular operation is so common that someone already wrote the mean function and it is included in base R. However, you will encounter situations in which the function does not already exist, so R permits you to write your own. A simple version of a function that computes the average can be defined like this:
```{r}
avg <- function(x){
  s <- sum(x)
  n <- length(x)
  s/n
}
```

Now avg is a function that computes the mean:
```{r}
x <- 1:100
identical(mean(x), avg(x))
```

Notice that variables defined inside a function are not saved in the workspace. So while we use s and n when we call avg, the values are created and changed only during the call. Here is an illustrative example:
```{r}
s <- 3
avg(1:10)
s
```
Note how s is still 3 after we call avg.  
In general, functions are objects, so we assign them to variable names with <-. The function function tells R you are about to define a function. The general form of a function definition looks like this:
```{r}
my_function <- function(VARIABLE_NAME){
  perform operations on VARIABLE_NAME and calculate VALUE
  VALUE
}
```
The functions you define can have multiple arguments as well as default values. For example, we can define a function that computes either the arithmetic or geometric average depending on a user defined variable like this:
```{r}
avg <- function(x, arithmetic = TRUE){
  n <- length(x)
  ifelse(arithmetic, sum(x)/n, prod(x)^(1/n))
}
```
We will learn more about how to create functions through experience as we face more complex tasks.
